Enforcing idempotency in neural networks

### 2025-01-30 paper.tex
Update with Jamie's comments taken into account2

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{multirow}
\usepackage{pythonhighlight}

@@

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2025}

@@

\begin{abstract}
    In this work, we propose a new architecture-agnostic method for training idempotent neural networks. An idempotent operator satisfies $f(x) = f(f(x))$, meaning it can be applied iteratively with no effect beyond the first application. Neural networks used in data transformation tasks, such as image generation and augmentation, represent non-linear idempotent projections. Using methods from perturbation theory we derive the recurrence relation ${\mathbf{K}' \leftarrow 3\mathbf{K}^2 - 2\mathbf{K}^3}$ for iteratively projecting a real-valued matrix $\mathbf{K}$ onto the manifold of idempotent matrices. Our analysis shows that for linear, single-layer MLP networks this projection 1) has idempotent fixed points, and 2) is attracting only around idempotent points. We give an extension to non-linear networks by considering our approach as a substitution of the gradient for the canonical loss function, achieving an architecture-agnostic training scheme. We provide experimental results for MLP- and CNN-based architectures with significant improvement in idempotent error over the canonical gradient-based approach. Finally, we demonstrate practical applications of the method as we train a generative network successfully using only a simple reconstruction loss paired with our method.
\end{abstract}

@@

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{mao-deblurring, ilesanmi-denoising, liu-gans}, and image generation \cite{liu-gans}, among others. While there is a variety of such transformation tasks, some of these can be considered \textit{intrinsically idempotent} in the sense that any function which carries out the transformation is necessarily an idempotent operator. An idempotent operation is one which can be applied iteratively with no effect beyond the first application. For instance, a sorting transformation is intrinsically idempotent since sorting an already sorted data structure is redundant and such a sorting is often deterministic. On the other hand, image transformations might not always be idempotent: rotating an image by 90$^{\circ}$ twice is not the same as rotating it once, for instance. As such, some transformations are intrinsically idempotent and admit \textit{only} idempotent solutions while other tasks are not and admit \textit{no} idempotent solutions. This work, however, is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

% What is an idempotent network even?
In this paper we are primarily concerned with networks ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$, where $\vtheta$ is a collection of weight parameters. The condition that $f_{\vtheta}$ is idempotent is
%
\begin{equation}
    f_{\vtheta}(\vx) = f_{\vtheta}(f_{\vtheta}(\vx)),
    \label{eq:idem}
\end{equation}
%
for all ${\vx \in \mathbb{R}^n}$. If ${f_{\vtheta}(\vx) = \vW \vx}$ (a single-layer, fully-connected network with no bias and the identity activation function) where ${\vW \in \mathbb{R}^{n \times n}}$ is the weight matrix, then condition (\ref{eq:idem}) reduces to the familiar notion from linear algebra where $\vW = \vW^2$ and eigenvalues of $\vW$ are either 0 or 1. Condition (\ref{eq:idem}) also gives the correct notion for non-linear networks acting as idempotent projections, and can be optimized using a simple mean-squared error loss
%
\begin{align}
    \mathcal{L}_\text{idem}(\vx) = \frac{1}{m} \sum_{i = 1}^m \left(f_{\vtheta}(f_{\vtheta}(\vx)) - f_{\vtheta}\big(\vx \big)\right)^2,
    \label{eq:idem-loss}
\end{align}
%
where ${\vx \in \mathbb{R}^{n}}$. As we show in Section \ref{sec:experiment}, minimizing this loss using canonical gradient descent can yield relatively poor improvement in the idempotent loss. Additionally, due to the higher-order application of $f_{\vtheta}$ the number of terms in the gradient ${\nabla_{\vtheta} \mathcal{L}_{\text{idem}}}$ grows exponentially in the number of layers, making the approach computationally expensive for certain architectures.

@@

Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. (\ref{eq:g}) as a modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks.

@@

We first define the central term \textit{near-idempotent}, used throughout:

@@

In the case when $n=1$ we consider ${\vD^2 \approx \bm{0}}$ and the expression ${(\vK')^2 - \vK' = \bm{0}}$ can be expanded and reduced by recursively applying the assumptions
%
\begin{align}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0},
\end{align}
%
for all $\vX,\vY,\vZ$ matrices. When $j\leq2$, \textit{i.e.}, the ansatz in Eq. (\ref{eq:ansatz}) is ${\vK' = \alpha_1 \vK + \alpha_2 \vK^2}$, there exists no solutions for $\alpha_i$. When $j=3$, however, there is exactly one solution when ${\alpha_1 = 0}$, ${\alpha_2 = 3}$ and ${\alpha_3 = -2}$, thus $g$ is as defined in Eq. (\ref{eq:g}). For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$ and is therefore generally less costly to compute.

@@

In general, we place no restrictions on the matrix ${\vK \in \mathbb{R}^{m \times m}}$. In particular, it might not be directly diagonalizable. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ such that
\begin{align}
    \vK = \vP \vJ \vP^{-1}
\end{align}
where $\vJ \in \mathbb{C}^{m \times m}$ is the Jordan normal form \cite{jordan-form} of ${\vK \in \mathbb{R}^{m \times m}}$. From this the dual problem,
\begin{align}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align}
can be constructed. The block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}
where $d_\lambda$ denotes the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Clearly, this system of equations is inconsistent when ${d_{\lambda} \geq 2}$, hence geometric multiplicity of every eigenvalue must also be exactly 1. Together this implies that $\vJ$ is diagonalizable for any fixed point $\vK$. Furthermore, the solutions which satisfy only Eq. (\ref{eq:cond1}) are:
%
\begin{align}
    \lambda \in \{0, 0.5, 1\}.
\end{align}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=0.99\textwidth]{./resources/fractal.pdf}
    \caption{10-time recursive application of $h(\lambda) = 3\lambda^2 - 2\lambda^3$ for each point on the complex plane. Black areas denote points converging onto $0$, while orange areas denote points converging onto $1$.}
    \label{fig:fractal}
\end{figure*}

Whilst this only technically applies in the linearized setting, we propose to also apply the method in non-linear settings using the recurrence relation
%
\begin{align}
    \vK' = \vK + \gamma(g(\vK) - \vK)
\end{align}
%
for $0 \leq \gamma \leq 1$. This has the effect of taking small $\gamma$-sized steps in the direction of $g$ at every time point.

@@

@@

% There is future work in uncovering the relationship between the suggested gradient from MB and that of OB

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=0.9\textwidth]{./resources/abs_err_b1234.png}
    \caption{Average of 10 runs of each algorithm for a variety of learning rates. Networks are randomly initialized and trained for $2\,500$ epochs. Runs which did not return a network with lower idempotent error than the initial value are discarded, and the average is over remaining runs. For networks B3 and B4, learning rates $>0.22$ and $>0.52$ respectively had no runs with improvement in error. For Modified Backpropagation on B1, some runs resulted in approximately $\bm{0}$ which, due to floating-point imprecision, results in the error spikes.}
    \label{fig:avg-abs-err}
\end{figure*}
%
\begin{figure*}[htbp]
    \centering
    \includegraphics[width=0.7\textwidth]{./resources/runs_err_b23.png}
    \caption{On networks B2 and B3, the average idempotent error across 10 runs for each learning rate is reported for each algorithm. Each column of graphs represents one algorithm. Modified Backpropagation achieves lower idempotent error at lower learning rates than Ordinary Backpropagation. The biggest relative improvement between algorithms occurs in the first $\sim500$ epochs.}
    \label{fig:avg-epoch-err}
\end{figure*}

\subsection{Application to Generative Networks}
\label{sec:experiment-gen}
%\textit{We replicate the results of IGN on MNIST and CelebA. Latent space analysis. Takeaway is that we show application to a U-net GAN architecture based on Conv layers.}
As mentioned, one of the motivating factors for actively enforcing idempotency during training is to apply it as a secondary optimization objective in conjunction with optimizing for a primary task. In this section we replicate the results of \citealt{shocher-ign} as we train a U-net style DCGAN architecture (see Appendix \ref{app:gen-training-scheme}) on the MNIST dataset. Let $\mathcal{D}$ denote the distribution of MNIST samples, while ${\mathcal{D}' = \mathcal{N}(0, 1)}$ from which noise is sampled. In this training scheme, the loss function being optimized is
%
\begin{align}
    \begin{split}
        \mathcal{L}{(\vtheta, \vtheta')} & = \lambda_r \mathcal{L}_{\mathrm{rec}}{(\vtheta)}                                                                          \\
                                         & + \lambda_i \mathcal{L}_{\mathrm{idem}}{(\vtheta, \vtheta')} + \lambda_r \mathcal{L}_{\mathrm{tight}}{(\vtheta, \vtheta')}
    \end{split}
    \label{eq:gan-loss}
\end{align}
%
where $\vtheta'$ is a copy of $\vtheta$ at each time step where $\vtheta'$ is detached from the computational graph. To see why this matters, consider $(\vx, \vy^*) \sim \mathcal{D}$ and $\vz \sim \mathcal{D}'$ and the individual loss components:
%
\begin{align}
    \mathcal{L}_{\mathrm{rec}}{((\vx, \vy^*); \vtheta)}    & = \Vert \vy^* - f_{\vtheta}(\vx) \Vert_1                            \\
    \mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}  & = \Vert f_{\vtheta'}(f_{\vtheta}(\vz)) - f_{\vtheta}(\vz) \Vert_1   \\
    \mathcal{L}_{\mathrm{tight}}{(\vz; \vtheta, \vtheta')} & = -\Vert f_{\vtheta}(f_{\vtheta'}(\vz)) - f_{\vtheta'}(\vz) \Vert_1
\end{align}
%
For instance, the quantity $\pd{\mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}}{\vtheta}$ is only affected by the inner application of $f$ above due to $\vtheta'$ being detached from the graph. The relationship between loss components $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$ is adversarial in nature.

@@

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=0.7\textwidth]{./resources/modified_0-1_0-1_rand_noise_mapping.png}
    \caption{Uncurated generations of the U-net style DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. In the top row are samples drawn from a random distribution with mean $0$ and variance $1$, whist the second and third rows represent first and second application of the network to these samples, respectively.}
    \label{fig:gen-mnist}
\end{figure*}

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{./resources/modified_0-1_0-1_rand_latent_space.png}
    \caption{Latent space manipulation for the U-net style DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. Samples $\mathbf{A}$ and $\mathbf{B}$ are selected randomly while remaining samples are linear combinations of these. We give the first and second application of the model.}
    \label{fig:latent-mnist}
\end{figure*}

@@

\newpage
\section{Generative samples}
\label{app:gen-samples}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.65\textwidth]{./resources/modified_0-1_0-1_rand_samples.png}
    \caption{Uncurated generations of U-DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. All samples are drawn from a random distribution with mean $0$ and variance $1$, and the result of applying the network is shown.}
    \label{fig:big-gen-mnist}
\end{figure}

>>>>

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{caption}
\usepackage{booktabs} % for professional tables
\usepackage{multirow}
\usepackage{pythonhighlight}
%\usepackage{placeins}

@@

% Use the following line for the initial blind version submitted for review:
%\usepackage{icml2025}

@@

\begin{abstract}
    In this work, we propose a new architecture-agnostic method for training idempotent neural networks. An idempotent operator satisfies $f(x) = f(f(x))$, meaning it can be applied iteratively with no effect beyond the first application. Some neural networks used in data transformation tasks, such as image generation and augmentation, can represent non-linear idempotent projections. Using methods from perturbation theory we derive the recurrence relation ${\mathbf{K}' \leftarrow 3\mathbf{K}^2 - 2\mathbf{K}^3}$ for iteratively projecting a real-valued matrix $\mathbf{K}$ onto the manifold of idempotent matrices. Our analysis shows that for linear, single-layer MLP networks this projection 1) has idempotent fixed points, and 2) is attracting only around idempotent points. We give an extension to non-linear networks by considering our approach as a substitution of the gradient for the canonical loss function, achieving an architecture-agnostic training scheme. We provide experimental results for MLP- and CNN-based architectures with significant improvement in idempotent error over the canonical gradient-based approach. Finally, we demonstrate practical applications of the method as we train a generative network successfully using only a simple reconstruction loss paired with our method.
\end{abstract}

@@

% Wider context -> Idempotent networks
Using neural networks as data augmentation tools is becoming more widespread in areas such as signal processing and generative artificial intelligence. In particular, networks of the form ${f: X \to X}$, mapping data within the same space $X$, are frequently used in image augmentation \cite{lu-image-aug}, video generation \cite{ma-compression, liu-gans}, sorting algorithms \cite{tambouratzis-sorting}, compression algorithms \cite{namphol-compression, liu-gans}, image denoising \cite{mao-deblurring, ilesanmi-denoising, liu-gans}, and image generation \cite{liu-gans}, among others.

Some data transformation tasks admit \textit{only} idempotent solutions (\textit{e.g.}, sorting), whilst other tasks admit \textit{no} idempotent solutions (\textit{e.g.}, rotating an image by 90$^{\circ}$). This work is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

%While there is a variety of such transformation tasks, some of these can be considered \textit{intrinsically idempotent} in the sense that any function which carries out the transformation is necessarily an idempotent operator. An idempotent operation is one which can be applied iteratively with no effect beyond the first application. For instance, a sorting transformation is intrinsically idempotent since sorting an already sorted data structure is redundant and such a sorting is often deterministic. On the other hand, image transformations might not always be idempotent: rotating an image by 90$^{\circ}$ twice is not the same as rotating it once, for instance. As such, some transformations are intrinsically idempotent and admit \textit{only} idempotent solutions while other tasks are not and admit \textit{no} idempotent solutions. This work, however, is concerned with a class of data transformation tasks which has both idempotent and non-idempotent solutions and where idempotency might be a desirable property. For example, in Section \ref{sec:experiment} we study idempotency in generative networks where it is the formal requirement of one-step inference, but also denoising and image augmentation (\textit{e.g.}, application of effect-filters) are examples of tasks where idempotent solutions may be desirable \cite{mao-deblurring,liu-gans}. Since solutions are not intrinsically idempotent in this class, we explore actively enforcing idempotency as a component of the loss function used in training.

% What is an idempotent network even?
In this paper we are primarily concerned with networks ${f_{\vtheta}: \mathbb{R}^n \to \mathbb{R}^n}$, where $\vtheta$ is a collection of weight parameters. The condition that $f_{\vtheta}$ is idempotent is the following, for all ${\vx \in \mathbb{R}^n}$:
%
\begin{equation}
    f_{\vtheta}(\vx) = f_{\vtheta}(f_{\vtheta}(\vx)).
    \label{eq:idem}
\end{equation}
%
If ${f_{\vtheta}(\vx) = \vW \vx}$ (a single-layer, fully-connected network with no bias and the identity activation function) where ${\vW \in \mathbb{R}^{n \times n}}$ is the weight matrix, then condition (\ref{eq:idem}) reduces to the familiar notion from linear algebra where $\vW = \vW^2$ and eigenvalues of $\vW$ are either 0 or 1. Condition (\ref{eq:idem}) also gives the correct notion for non-linear networks acting as idempotent projections, and can be optimized using a simple mean-squared error loss, where ${\vx \in \mathbb{R}^{n}}$:
%
\begin{align}
    \mathcal{L}_\text{idem}(\vx) = \frac{1}{m} \sum_{i = 1}^m \left(f_{\vtheta}(f_{\vtheta}(\vx)) - f_{\vtheta}\big(\vx \big)\right)^2.
    \label{eq:idem-loss}
\end{align}
%
As we show in Section \ref{sec:experiment}, minimizing this loss using canonical gradient descent can yield relatively poor improvement in the idempotent loss. Additionally, due to the higher-order application of $f_{\vtheta}$ the number of terms in the gradient ${\nabla_{\vtheta} \mathcal{L}_{\text{idem}}}$ grows exponentially in the number of layers, making the approach computationally expensive for certain architectures.

@@

Although we assume $\vK$ is close to idempotent, we show that in practice $g$ can be used to derive matrices which are within machine precision of perfect idempotency even when the input matrix $\vK$ is relatively far from idempotent. At a high level, this process is based on a recurrence relation ${\vK' = \vK + \gamma(g(\vK) - \vK)}$, taking small $\gamma$-sized steps in the direction of $g(\vK)$. While this recurrence relation derives idempotent matrices -- and can therefore be used to train single-layer networks with identity activations to be idempotent -- we also give a more general application of Eq. (\ref{eq:g}) as a modification of the backpropagation algorithm, yielding an architecture agnostic and efficient algorithm for finding idempotent networks. As we will see, this modification in general not only leads to higher idempotent error reduction but also seems to explore the loss landscape differently from the canonical approach.

@@

We first define the term \textit{near-idempotent} used throughout:

@@

In the case when $n=1$ we consider ${\vD^2 \approx \bm{0}}$ and the expression ${(\vK')^2 - \vK' = \bm{0}}$ can be expanded and reduced by recursively applying the following assumptions, for all $\vX,\vY,\vZ$ matrices:
%
\begin{align}
    \vD^2 \approx \bm{0}, \quad \vP^2 = \vP, \quad \vX \vD \vY \vD \vZ \approx \bm{0}.
\end{align}
%
When $j\leq2$, \textit{i.e.}, there exists no solutions for $\alpha_i$. When $j=3$, however, there is exactly one solution when ${\alpha_1 = 0}$, ${\alpha_2 = 3}$ and ${\alpha_3 = -2}$, thus $g$ is as defined in Eq. (\ref{eq:g}). For $j>3$ there exists families of solutions (see Appendix \ref{app:solutions}), but we consider primarily the case when $j=3$ as this requires fewer higher-order terms of $\vK$ and is therefore generally less costly to compute.

@@

In general, we place no restrictions on the matrix ${\vK \in \mathbb{R}^{m \times m}}$. In particular, it might not be directly diagonalizable. It is well known, however, that for every square matrix $\vK$ there exists an invertible matrix $\vP$ and a Jordan normal form \cite{jordan-form} $\vJ \in \mathbb{C}^{m \times m}$ of ${\vK \in \mathbb{R}^{m \times m}}$ such that ${\vK = \vP \vJ \vP^{-1}}$. From this the dual problem,
\begin{align}
    \vJ = 3 \vJ^2 - 2 \vJ^3,
\end{align}
can be constructed. Let $d_\lambda$ denotes the algebraic multiplicity of eigenvalue $\lambda$ of $\vK$. Then, the block-diagonal structure of $\vJ$ imposes up to four equations per eigenvalue of $\vK$ (see Appendix \ref{app:jordan}):
\begin{align}
    \lambda & = 3\lambda^2 - 2\lambda^3 & \label{eq:cond1}                                   \\
    1       & = 6\lambda - 6\lambda^2   & \text{Only when $d_\lambda\geq2$.}\label{eq:cond2} \\
    0       & = 3 - 6\lambda            & \text{Only when $d_\lambda\geq3$.}\label{eq:cond3} \\
    0       & = 0 - 2                   & \text{Only when $d_\lambda\geq4$.}\label{eq:cond4}
\end{align}

Clearly, this system of equations is inconsistent when ${d_{\lambda} \geq 2}$, hence geometric multiplicity of every eigenvalue must also be exactly 1. Together this implies that $\vJ$ is diagonalizable for any fixed point $\vK$. Furthermore, the solutions which satisfy only Eq. (\ref{eq:cond1}) are:
%
\begin{align}
    \lambda \in \{0, 0.5, 1\}.
\end{align}

@@

\begin{figure*}[!htp]
    \centering
    \includegraphics[width=\textwidth]{./resources/fractal.pdf}
    \caption{10-time recursive application of $h(\lambda) = 3\lambda^2 - 2\lambda^3$ for each point on the complex plane. Black areas denote points converging onto $0$, while orange areas denote points converging onto $1$.}
    \label{fig:fractal}
\end{figure*}

Whilst this only technically applies in the linearized setting, we propose to also apply the method in non-linear settings using the following recurrence relation, for $0 \leq \gamma \leq 1$:
%
\begin{align}
    \vK' = \vK + \gamma(g(\vK) - \vK).
\end{align}
%
This has the effect of taking small $\gamma$-sized steps in the direction of $g$ at every time point.

@@

%\begin{figure*}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./resources/abs_err_b1234.png}
%    \caption{Average of 10 runs of each algorithm for a variety of learning rates. Networks are randomly %initialized and trained for $2\,500$ epochs. Runs which did not return a network with lower idempotent %error than the initial value are discarded, and the average is over remaining runs. For networks B3 %and B4, learning rates $>0.22$ and $>0.52$ respectively had no runs with improvement in error. For %Modified Backpropagation on B1, some runs resulted in approximately $\bm{0}$ which, due to %floating-point imprecision, results in the error spikes.}
%    \label{fig:avg-abs-err}
%\end{figure*}
%%
%\begin{figure*}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./resources/runs_err_b23.png}
%    \caption{On networks B2 and B3, the average idempotent error across 10 runs for each learning rate is %reported for each algorithm. Each column of graphs represents one algorithm. Modified Backpropagation %achieves lower idempotent error at lower learning rates than Ordinary Backpropagation. The biggest %relative improvement between algorithms occurs in the first $\sim500$ epochs.}
%    \label{fig:avg-epoch-err}
%\end{figure*}

%\begin{figure*}[!t]
%    \centering
%    \subfigure[Average of 10 runs of each algorithm for a variety of learning rates. Networks are randomly %initialized and trained for $2\,500$ epochs. Runs which did not return a network with lower idempotent %error than the initial value are discarded, and the average is over remaining runs. For networks B3 %and B4, learning rates $>0.22$ and $>0.52$ respectively had no runs with improvement in error. For %Modified Backpropagation on B1, some runs resulted in approximately $\bm{0}$ which, due to %floating-point imprecision, results in the error spikes.]{
%        \includegraphics[width=0.9\linewidth]{./resources/abs_err_b1234.png}
%        \label{fig:avg-abs-err}
%    }
%    \subfigure[On networks B2 and B3, the average idempotent error across 10 runs for each learning rate %is reported for each algorithm. Each column of graphs represents one algorithm. Modified %Backpropagation achieves lower idempotent error at lower learning rates than Ordinary Backpropagation. %The biggest relative improvement between algorithms occurs in the first $\sim500$ epochs.]{
%        \includegraphics[width=0.8\linewidth]{./resources/runs_err_b23.png}
%        \label{fig:avg-epoch-err}
%    }
%\end{figure*}

\begin{figure*}[!t]
    \centering
    \captionbox{Average of 10 runs of each algorithm for a variety of learning rates. Networks are randomly initialized and trained for $2\,500$ epochs. Runs which did not return a network with lower idempotent error than the initial value are discarded, and the average is over remaining runs. For networks B3 and B4, learning rates $>0.22$ and $>0.52$ respectively had no runs with improvement in error. For Modified Backpropagation on B1, some runs resulted in approximately $\bm{0}$ which, due to floating-point imprecision, results in the error spikes.\label{fig:avg-abs-err}}{
        \includegraphics[width=0.9\linewidth]{./resources/abs_err_b1234.png}

    }
    \captionbox{On networks B2 and B3, the average idempotent error across 10 runs for each learning rate is reported for each algorithm. Each column of graphs represents one algorithm. Modified Backpropagation achieves lower idempotent error at lower learning rates than Ordinary Backpropagation. The biggest relative improvement between algorithms occurs in the first $\sim500$ epochs.\label{fig:avg-epoch-err}}{
        \includegraphics[width=0.8\linewidth]{./resources/runs_err_b23.png}

    }
\end{figure*}

@@

\begin{figure*}[!b]
    \centering
    \captionbox{Uncurated generations of the U-net style DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. In the top row are samples drawn from a random distribution with mean $0$ and variance $1$, whist the second and third rows represent first and second application of the network to these samples, respectively.\label{fig:gen-mnist}}{
        \includegraphics[width=0.7\textwidth]{./resources/modified_0-1_0-1_rand_noise_mapping.png}
    }
    \captionbox{Latent space manipulation for the U-net style DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. Samples $\mathbf{A}$ and $\mathbf{B}$ are selected randomly while remaining samples are linear combinations of these. We give the first and second application of the model.\label{fig:latent-mnist}}{
        \includegraphics[width=0.8\textwidth]{./resources/modified_0-1_0-1_rand_latent_space.png}
    }
\end{figure*}

% There is future work in uncovering the relationship between the suggested gradient from MB and that of OB
\subsection{Application to Generative Networks}
\label{sec:experiment-gen}
%\textit{We replicate the results of IGN on MNIST and CelebA. Latent space analysis. Takeaway is that we show application to a U-net GAN architecture based on Conv layers.}
As mentioned, one of the motivating factors for actively enforcing idempotency during training is to apply it as a secondary optimization objective in conjunction with optimizing for a primary task. In this section we replicate the results of \citealt{shocher-ign} as we train a U-net style DCGAN architecture (see Appendix \ref{app:gen-training-scheme}) on the MNIST dataset. Let $\mathcal{D}$ denote the distribution of MNIST samples, while ${\mathcal{D}' = \mathcal{N}(0, 1)}$ from which noise is sampled. Let $\vtheta'$ is a copy of the trainable weights $\vtheta$ at each time step, where $\vtheta'$ is detached from the computational graph. In this training scheme, the loss function being optimized is
%
\begin{align}
    \begin{split}
        \mathcal{L}{(\vtheta, \vtheta')} & = \lambda_r \mathcal{L}_{\mathrm{rec}}{(\vtheta)}                                                                           \\
                                         & + \lambda_i \mathcal{L}_{\mathrm{idem}}{(\vtheta, \vtheta')} + \lambda_r \mathcal{L}_{\mathrm{tight}}{(\vtheta, \vtheta')}.
    \end{split}
    \label{eq:gan-loss}
\end{align}
%
To see why employing two copies of the weights is useful, consider $(\vx, \vy^*) \sim \mathcal{D}$ and $\vz \sim \mathcal{D}'$ and the individual loss components:
%
\begin{align}
    \mathcal{L}_{\mathrm{rec}}{((\vx, \vy^*); \vtheta)}    & = \Vert \vy^* - f_{\vtheta}(\vx) \Vert_1                            \\
    \mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}  & = \Vert f_{\vtheta'}(f_{\vtheta}(\vz)) - f_{\vtheta}(\vz) \Vert_1   \\
    \mathcal{L}_{\mathrm{tight}}{(\vz; \vtheta, \vtheta')} & = -\Vert f_{\vtheta}(f_{\vtheta'}(\vz)) - f_{\vtheta'}(\vz) \Vert_1
\end{align}
%
For instance, the quantity $\pd{\mathcal{L}_{\mathrm{idem}}{(\vz; \vtheta, \vtheta')}}{\vtheta}$ is only affected by the inner application of $f$ above due to $\vtheta'$ being detached from the graph. The relationship between loss components $\mathcal{L}_{\mathrm{idem}}$ and $\mathcal{L}_{\mathrm{tight}}$ is adversarial in nature.

@@

%\begin{figure*}[htbp]
%    \centering
%    \includegraphics[width=0.7\textwidth]{./resources/modified_0-1_0-1_rand_noise_mapping.png}
%    \caption{Uncurated generations of the U-net style DCGAN model trained on MNIST with Modified %Backpropagation for optimizing idempotent and tightness losses. In the top row are samples drawn from %a random distribution with mean $0$ and variance $1$, whist the second and third rows represent first %and second application of the network to these samples, respectively.}
%    \label{fig:gen-mnist}
%\end{figure*}
%
%
%\begin{figure*}[htbp]
%    \centering
%    \includegraphics[width=0.8\textwidth]{./resources/modified_0-1_0-1_rand_latent_space.png}
%    \caption{Latent space manipulation for the U-net style DCGAN model trained on MNIST with Modified %Backpropagation for optimizing idempotent and tightness losses. Samples $\mathbf{A}$ and $\mathbf{B}$ %are selected randomly while remaining samples are linear combinations of these. We give the first and %second application of the model.}
%    \label{fig:latent-mnist}
%\end{figure*}

@@

\newpage
\section{Generative samples}
\label{app:gen-samples}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.65\textwidth]{./resources/modified_0-1_0-1_rand_samples.png}
    \caption{Uncurated generations of U-net style DCGAN model trained on MNIST with Modified Backpropagation for optimizing idempotent and tightness losses. All samples are drawn from a random distribution with mean $0$ and variance $1$, and the result of applying the network is shown.}
    \label{fig:big-gen-mnist}
\end{figure}
